---
title: "HW10 - Ordination"
author: "Evan Batzer"
date: "December 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

##### Dune

dune is a data frame of observations of species abundances at different sites. 

##### Dune.env

dune.env is a data frame of the following variables:

A1: a numeric vector of thickness of soil A1 horizon.

Moisture: an ordered factor with levels: 1 < 2 < 4 < 5.

Management: a factor with levels: BF (Biological farming), HF (Hobby farming), NM (Nature Conservation Management), and SF (Standard Farming).

Use: an ordered factor of land-use with levels: Hayfield < Haypastu < Pasture.

Manure: an ordered factor with levels: 0 < 1 < 2 < 3 < 4.

________________________________________________________________________________

### Questions

#### 1. Call and provide a basic overview of the `dune` datasets from the `vegan` package. You will need to retrieve both the community data, `dune`, and the associated environmental data, `dune.env`.  

*Answer the following:*   
1. How many species and samples are included in the community dataset?
2. How many environmental attributes were measured in the environmental dataset?
3. What is the average abundance of each species across all samples? Sort these in descending order.

#### Answer:

```{r}
library(vegan)

data(dune)
data(dune.env)

dim(dune) # 20 samples, 30 species
dim(dune.env) # 20 samples, 5 attributes

sort(colMeans(dune), decreasing = TRUE)
```

#### 2. Produce a heatmap of the abundances of species across samples

#### Answer:

```{r}
tabasco(dune)
```

#### 3. What do you notice about species distributions in this dataset? Is this dataset well-suited to common ordination approaches in ecology?

#### Answer:

#### 4. Visualize these samples using non-metric multidimensional scaling (NMDS). Use the Bray-Curtis (`bray`) dissimilarity metric. Color sites by the "Use" environmental factor.

__Note:__ To set custom colors, you will need to first call `ordiplot(nmds_object, type = "n")`, then add points from the nmds with points()

```{r}
my_nmds <- metaMDS(dune, distance = "bray")

ordiplot(my_nmds, type = "n")
points(my_nmds, "sites", col = dune.env$Use)
```

