---
title: "ModelContrasts"
author: "Evan Batzer"
date: "February 24, 2020"
output: html_document
---

# Linear hypothesis tests

Linear hypothesis tests involve a *linear* combination of the components of a statistical model, usually to generate the estimated mean value of our response given some set of predictors. 

In most cases, these estimated means will be:

* The response of a group in an ANOVA
* The response at some value of a predictor in regression

#### What is meant by *linear combination*?

Mathematically speaking, linear operations used in hypothesis testing are either:

* Addition (X + Y)
* Scalar multiplication (cX)

To understand how we construct hypothesis tests using these components, we need to think about how most models are constructed.

Let's consider a one-way ANOVA, comparing the estimated responses of three groups. I'll use the `iris` dataset here, for convenience.

```{r, echo = FALSE, message = FALSE}
library(tidyverse); library(emmeans)

data(iris)

iris %>% ggplot(aes(x = Species,
                    y = Sepal.Width,
                    color = Species)) +
  geom_boxplot() +
  guides(color = FALSE) +
  ggtitle("Sepal Length by Species")
```

The ANOVA model used in this case could be written as `lm(Sepal.Length ~ Species, data = iris)`

Given how R uses factor coding, the fitted model will consist of three coefficients:

* $\beta_0$, the estimated Sepal Length of the first group, "setosa"
* $\beta_1$, the estimated difference in Sepal Length between the second group, "versicolor", and the first
* $\beta_2$, the estimated difference in Sepal Length between the third group, "virginica", and the first

Or, written mathematically:

$(\hat Y_{setosa}) = \beta_0$  
$(\hat Y_{versicolor}) = \beta_0 + \beta_1$  
$(\hat Y_{virginica}) = \beta_0 + \beta_2$  

### What tests can we run with this data?

The most natural test to start with would be asking whether the groups differ from one another. Mathematically, this statement could be expressed as whether $\hat Y_{species 1} - \hat Y_{species 1} = 0$

Based on how this model is constructed, we express this difference using coefficients:

$(\hat Y_{setosa}) - (\hat Y_{versicolor}) = (\beta_0) - (\beta_0 + \beta_1) = \beta_1$  
$(\hat Y_{setosa}) - (\hat Y_{virginica}) = (\beta_0) - (\beta_0 + \beta_2) = \beta_2$  
$(\hat Y_{versicolor}) - (\hat Y_{virginica}) = (\beta_0) - (\beta_0 + \beta_1) = \beta_1 - \beta_2$  

If we were to express these $\beta$ coefficients as a vector $\vec{\beta} = (\beta_0, \beta_1, \beta_2)$, we could write these tests as the following:  

$(\hat Y_{setosa}) - (\hat Y_{versicolor}) = (0,1,0) * \vec{\beta}$  
$(\hat Y_{setosa}) - (\hat Y_{virginica}) = (0,0,1) * \vec{\beta}$  
$(\hat Y_{versicolor}) - (\hat Y_{virginica} = (0,1,-1) * \vec{\beta}$  

Where each hypothesis is a linear combination of coefficients, either added or subtrated from one another with a constant multiplier (1).

### How do I run these tests?

By design, some tests are baked in to `lm` when using factor variables. The summary object will offer T-tests of whether each coefficient differs from zero, which can be used to determine whether the sepal width of the first group, setosa, differs from 0, and whether the other two groups differ from this first factor level.

```{r}
iris_anova <- lm(Sepal.Width ~ Species, data = iris)
summary(iris_anova)
```

But how do we test the third hypothesis, whether the second and third groups differ from one another? Enter `emmeans` -- a useful package to generate "estimated marginal means" that can be used to test hypothesis. By default, the `pairs()` function quickly compares means between groups in all possible combinations.

```{r}
# Call emmeans on the model object, with one or more sets of columns to generate means for
iris_emm <- emmeans(iris_anova, specs = "Species")

# All pairwise contrasts of the "Species" column
pairs(iris_emm)
```

But we could also do this in a more defined way, by specifying linear combinations of these estimated means, as shown before. The one twist is that `emmeans` would like to define the linear combinations by estimated means, so the contrasts will take a "1" or "-1" for each mean that is being compared:

```{r}
test1 <- c(1,-1,0) # mean of species 1 minus species 2
test2 <- c(1,0,-1) # mean of species 1 minus species 3
test3 <- c(0,1,-1) # mean of species 2 minus species 3

contrast(iris_emm, 
         method = list("1. Se - Ve" = test1, 
                       "2. Se - Vi" = test2, 
                       "3. Ve - Vi" = test3))
```

### More complicated tests

At times, we might have specific biological hypothesis for how estimated means differ from one another

```{r}
test4 <- c(2, 0, -1)

contrast(iris_emm, 
         method = list("4. 2xSe - Vi" = test4))

test5 <- c(0, 1, -1)

contrast(iris_emm, 
         method = list("5. Se - Vi" = test4),
         offset = - 3.5)
```

### P-value control

```{r}
pairs(iris_emm)
```

```{r}
contrast(iris_emm, 
         method = list("1. Se - Ve" = test1, 
                       "2. Se - Vi" = test2, 
                       "3. Ve - Vi" = test3),
         adjust = "bonferroni")
```


### A worked example
